---
title: "R Notebook - Figure 4 and Supplementary Figure 4"
output: html_document
---
This notebook covers plots in Figure 4 and Supplementary Figure 4.

```{r setup, include=FALSE}
    knitr::opts_knit$set(root.dir = normalizePath('/Users/leppam/dkfz/venetoclax/manuscripts/code_github/')) 
```

Required packages and directories.

```{r, message=F}
library(data.table)
library(dplyr)
library(tibble)
library(stringr)
library(ggplot2)
library(ggrepel)
library(ggnewscale)
library(scales)
library(cowplot)
library(decoupleR)
library(ggalluvial)
library(ComplexHeatmap)

input_dir <- "./input/"
output_dir <- "./output/"

# Color script
source("./scripts/colors.R")
```

Read in data.

```{r}
# Stuart et al. healthy BM CITE-seq data
plot_stuart_sc_bm <- fread(paste0(input_dir, 'stuart_sc_bm_df.tsv'))

# Bulk projections
bulk_projections_df <- fread(paste0(input_dir, 'bulk_projections_df.tsv'))

# MATURE samples with CD64+CD11b+ cells > 40%
mature.df <- readxl::read_excel(paste0(input_dir, 'matures_to_plot.xlsx'))

# Clinical metadata of FACS sorted populations
metadata_populations <- fread(paste0(input_dir, 'metadata_populations.tsv'))

```

# Plot Figure 4A

Project classical immature GPR56+ LSCs and non-classical MoDe-LSCs onto a CITE-seq reference map of healthy hematopoiesis (Stuart et al.).

```{r}

# Combine bulk projections with Stuart et al. BM
bulk_projections_stuart_sc_bm <- bind_rows(bulk_projections_df, plot_stuart_sc_bm)

# Get cell type coordinates for labels
label_coord <- plot_stuart_sc_bm %>%
  group_by(celltype.l2) %>%
  summarise(refUMAP_1 = mean(refUMAP_1),
            refUMAP_2 = mean(refUMAP_2)) %>%
  ungroup() %>%
  filter(celltype.l2 %in% myeloid.cells)


title <- 'Immature GPR56+ LSCs & MoDe-LSCs'

p_lsc <- bulk_projections_stuart_sc_bm %>% 
  filter(celltype.l2 %in% c(myeloid.cells, 'LSC','MATURE')) %>%
  ggplot(aes(refUMAP_1, refUMAP_2)) +
  geom_point(data = . %>% filter(batch == "bmcite"), aes(color = celltype.l2), size = 0.2, alpha = 0.2) +
  scale_color_manual(values = lighten(col.myeloid, 0.2),
                     guide = 'none') +
  guides(color = 'none') +
  # Start a new scale
  ggnewscale::new_scale_fill() +
  geom_point(data = . %>% filter(batch == "combined" & celltype.l2 == 'LSC'),
             aes(fill = predicted.celltype), stroke = 0.25,
             size = 1, alpha = 0.8, shape = 21) +
  scale_fill_manual(values = darken(col.myeloid, 0.1),
                    labels = paste0(myeloid.cells, '-like')) +
  guides(fill = guide_legend(override.aes = list(shape = 21, size = 2), title = "Classical LSCs")) +
  # Start a new scale
  ggnewscale::new_scale_fill() +
  # Plot mature samples with mature >40%
  geom_point(data = . %>% filter(batch == "combined" & `Mono.LSCs` %in%  c('yes') & Cell %in% mature.df$Sample),
             aes(fill = `Mono.LSCs`),
             size = 1, alpha = 1, stroke = 0.25, shape = 21) +
  scale_fill_manual(values = col.MonoLSC,
                    labels = 'MoDe-LSC') +
  # Start a new scale
  ggnewscale::new_scale_color() +
  # Add cell type labels
  geom_text_repel(data = label_coord,
                  aes(refUMAP_1, refUMAP_2, label = celltype.l2, color = celltype.l2),
                  size = 2.5,
                  fontface = 2,
                  force = 5,
                  force_pull = 0.5,
                  max.overlaps = Inf,
                  box.padding = 0.8,
                  point.padding = 0.5,
                  direction = "both", 
                  segment.color = "black",
                  segment.size = 0.25) +
  scale_color_manual(values = darken(col.myeloid, 0.2), guide = NULL) +
  guides(fill = guide_legend(override.aes = list(shape = 21, size = 2), title = "Non-classical LSCs")) +
  ggtitle(paste(title)) +
  theme_classic() + 
  theme(legend.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=14),
        axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.ticks=element_blank()
  ) +
  coord_fixed() +
  labs(title = title,
       x = "UMAP 1",
       y = "UMAP 2")

p_lsc

save_plot(paste0(output_dir, 'Projection_LSCs.pdf'), p_lsc, base_height = 5, base_width = 6)

```

# Figure 4C

Plot the relative transcription factor expression of different LSC-types. Compare always against LMPP-like LSCs. 

```{r, fig.height = 3.5, fig.width = 15}
res_LMPP_MoDe <- fread(paste0(input_dir, 'DEG_LMPP_MoDe.tsv'))
res_LMPP_HSC <- fread(paste0(input_dir, 'DEG_LMPP_HSC.tsv'))
res_LMPP_MEP <- fread(paste0(input_dir, 'DEG_LMPP_MEP.tsv'))

# Get TFs using dorothea

# Retrieve all regulons from human (confidence levels A, B and C)
tf_genes <- decoupleR::get_dorothea(organism = "human", levels = c('A', 'B', 'C')) %>% pull(source) %>% unique()
length(tf_genes)

# Annotate TFs
res_LMPP_MoDe <- res_LMPP_MoDe %>%
  mutate(Comparison = 'LMPP vs. MoDe') %>%
  mutate(TF = case_when(SYMBOL %in% tf_genes ~ 'yes',
                        TRUE ~ 'no')) %>%
  # keep only TFs
  filter(TF == 'yes')

res_LMPP_HSC <- res_LMPP_HSC %>%
  mutate(Comparison = 'LMPP vs. HSC') %>%
  mutate(TF = case_when(SYMBOL %in% tf_genes ~ 'yes',
                        TRUE ~ 'no')) %>%
  # keep only TFs
  filter(TF == 'yes')

res_LMPP_MEP <- res_LMPP_MEP %>%
  mutate(Comparison = 'LMPP vs. MEP') %>%
  mutate(TF = case_when(SYMBOL %in% tf_genes ~ 'yes',
                        TRUE ~ 'no')) %>%
  # keep only TFs
  filter(TF == 'yes')

# All to LMPP
if(all(colnames(res_LMPP_HSC) == colnames(res_LMPP_MEP) & colnames(res_LMPP_HSC) == colnames(res_LMPP_MoDe)) ) {
  df.all <- bind_rows(res_LMPP_HSC, res_LMPP_MEP, res_LMPP_MoDe)
}

df.all <- df.all %>%
  mutate(Comparison = factor(Comparison, levels = c('LMPP vs. HSC', 'LMPP vs. MEP', 'LMPP vs. MoDe')))

# Bubble plot
p_thresh <- 0.1

# Define the order of comparisons
comp_levels <- c('LMPP vs. HSC', 'LMPP vs. MEP', 'LMPP vs. MoDe')

p_bubble <- df.all %>%
  filter(padj < 0.1 & abs(log2FoldChange) > 1.5) %>%
  group_by(SYMBOL) %>%
  mutate(
    # Assign a power-of-10 weight based on the Comparison level
    symbol_weight = sum(10^(4 - as.numeric(factor(Comparison, levels = comp_levels))))
  ) %>%
  ungroup() %>%
  # Reorder SYMBOL by this weight, then by log2FC to break ties
  mutate(SYMBOL = reorder(SYMBOL, symbol_weight + (log2FoldChange/100))) %>%
  
  ggplot(aes(x = Comparison, y = SYMBOL)) +
  geom_point(aes(fill = log2FoldChange, size = -log10(padj)),
             shape = 21) +
  scale_fill_gradient2(
    high = "#832424", low = "#3A3A98", mid = 'white', 
    midpoint = 0, oob = squish
  ) +
  scale_size_continuous(range = c(2, 8), name = "-log10(p-adj)") +
  labs(
    title = 'TF expression in LSC subtypes',
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    panel.border = element_rect(colour = "black", fill=NA, linewidth=0.25)
  ) +
  coord_flip()

p_bubble

save_plot(paste0(output_dir, 'TF_expression.pdf'), p_bubble, base_height = 3, base_width = 13)

```

# Supplementary Figure 4A

Project publicly available RNA-seq data of CD34/CD38 AML fractions from Zeng et al. (Nature Medicine 2022) onto a healthy BM reference (Stuart et al.). Compare the CD34/CD38 fractions with our GPR56+ LSCs and MoDe-LSCs projected onto the same reference. 

```{r}

# Zeng et al cell type predictions
zeng_df <- fread(file = paste0(input_dir, 'zeng_fractions_mapped.tsv'))

# VV cell type predictions
bulk_projections_df <- fread(paste0(input_dir, 'bulk_projections_df.tsv'))

# Keep only sorted population generated for this paper (remove healthy cells part of the df (SeuratData: "bmcite"))
df.vv <- bulk_projections_df %>%
  rename(Sample = Cell,
         Population = celltype.l2,
         Stage = disease_stage)

# Data wrangling 
keep.cols <- c('Sample','Population','Stage','predicted.celltype','Mono.LSCs')

# Keep only LSCs
df.vv.lsc <- df.vv %>%
  filter(Population %in% c('LSC','MATURE')) %>%
  dplyr::select(all_of(keep.cols)) %>%
  # if MoDe, keep MoDe-LSCs
  mutate(priority = case_when(grepl("yes", Mono.LSCs) & Population == 'MATURE' ~ 0,
                              Population == 'LSC' ~ 1,
                              TRUE ~ 2)) %>%  # lower = better
  group_by(Sample) %>%
  arrange(priority, row_number()) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  # Remove samples that only have a non-MoDe Mature
  filter(priority != 2) %>%
  # name the cohort
  mutate(cohort = 'Current data') %>%
  # add FACS population
  mutate(population = case_when(Population == 'LSC' ~ 'GPR56+ LSC',
                                Population == 'MATURE' ~ 'MoDe-LSC',
                                TRUE ~ NA)) %>%
  mutate(population = factor(population, levels = c('GPR56+ LSC','MoDe-LSC'))) %>%
  dplyr::select('Sample','population', 'predicted.celltype','cohort')

df.zeng.lsc <- zeng_df %>%
  # name the cohort
  mutate(cohort = 'Zeng et al.') %>%
  mutate(population = factor(population, levels = c('CD34+CD38-', 'CD34+CD38+', 'CD34-CD38-', 'CD34-CD38+'))) %>%
  rename(Sample = 'cell') %>%
  dplyr::select('Sample','population', 'predicted.celltype','cohort')

df.AML <- bind_rows(df.vv.lsc, df.zeng.lsc) %>%
  mutate(predicted.celltype = factor(predicted.celltype, levels = names(col.celltype)))

# Compare cell type changes between predicted cell type and facs cell type
df.alluvial <- df.AML %>%
  group_by(cohort) %>%
  count(predicted.celltype, population, name = "Count") %>%
  group_by(cohort, population) %>%
  mutate(Fraction = Count/sum(Count))

# Plot alluvial plots
p_alluvial <- df.alluvial %>%
  ggplot(aes(y = Count, axis1 = population, axis2 = predicted.celltype)) +
  ggalluvial::geom_alluvium(aes(fill = predicted.celltype), width = 1/12) +
  ggalluvial::geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum)), size = 2) +
  scale_fill_manual(values = col.celltype) +
  facet_wrap(vars(cohort), scales = 'free_y', ncol = 2) +
  scale_x_discrete(limits = c("population", "predicted.celltype"),
                   labels = c("FACS population", "BM ref. Stuart et al."), expand = c(.2, .2)) +
  labs(title = 'FACS to cell type predictions') +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, vjust = 1))

p_alluvial

save_plot(paste0(output_dir, 'FACS_to_celltype_predictions.pdf'), p_alluvial, ncol = 2, base_height = 5, base_width = 3)

```
# Supplementary Figure 4I

Plot oncoprints of recurrent mutations and cytogenetics grouped by LSC-type for diagnosis and salvage patient cohort.
Use FACS sorted population metadata that contains clinical and mutational information of the patient.

```{r}

# Keep only classical LSCs
meta.lsc <- metadata_populations %>%
  filter(Stage %in% c('Diagnosis','Salvage')) %>%
  filter(Population %in% c('LSC','MATURE')) %>%
  # Keep classical and non-classical LSCs
  filter( (Population == 'LSC' & (Mono.LSCs_engraftment_or_lamp5 == '' | Mono.LSCs_engraftment_or_lamp5 == 'no')) | 
          (Population == 'MATURE' & Mono.LSCs_engraftment_or_lamp5 == 'yes') ) %>%
  distinct(Sample, .keep_all = TRUE) %>%
  # Create new cell type label
  mutate(LSC_type = case_when(Mono.LSCs_engraftment_or_lamp5 == 'yes' ~ 'MoDe',
                              TRUE ~ predicted.celltype)) %>%
  mutate(LSC_type = factor(LSC_type, levels = c('HSC', 'LMPP', 'MEP','MoDe'))) %>%
  filter(!is.na(LSC_type))

# Patterns to group chromosomal alterations
meta.lsc$`-5/del(5q)` <- ifelse(str_detect(meta.lsc$Karyotype, paste(c("del\\(5\\)\\(q", "del5q", "\\-5"), collapse = '|')) == TRUE, 'loss', '')
meta.lsc$`-7/del(7q)` <- ifelse(str_detect(meta.lsc$Karyotype, paste(c("del\\(7\\)\\(q", "del7q", "\\-7"), collapse = '|')) == TRUE, 'loss', '')
meta.lsc$`-17/del(17p)` <- ifelse(str_detect(meta.lsc$Karyotype, paste(c("del\\(17\\)\\(p", "del17p", "\\-17"), collapse = '|')) == TRUE, 'loss', '')
meta.lsc$`+8/dup(8q)` <- ifelse(str_detect(meta.lsc$Karyotype, paste(c("dup\\(8\\)\\(q", "add\\(8\\)\\(q", "Trisomie 8q21", "\\+8"), collapse = '|')) == TRUE, 'gain', '')

mat <- meta.lsc %>% 
  select(Sample, KMT2A.re:c.Kit, `-5/del(5q)`:`+8/dup(8q)`) %>% 
  distinct(Sample, .keep_all = TRUE) %>%
  # Remove NAs
  filter(TP53 != 'N/A') %>%
  # Change yes/no to binary
  mutate_at(vars(TP53:c.Kit), funs(ifelse(. == 'yes', "snv", ""))) %>%
  mutate_at(vars(KMT2A.re:Complex), funs(ifelse(. == 'yes', "other", ""))) %>%
  rename('KMT2A-re' = KMT2A.re, 'IDH1/2' = IDH1.2, 'Splicing' = SF) %>%
  column_to_rownames('Sample') %>%
  as.matrix()

# In order to clearly visualize the data, genes for which less than 4% patients have mutations (l2) and patients which have mutations in less than 1% of the genes (l1) are removed.
l1 = apply(mat, 1, function(x) sum(!grepl("^\\s*$", x))/length(x) >= 0.01)
l2 = apply(mat, 2, function(x) sum(!grepl("^\\s*$", x))/length(x) >= 0.04)

mat.flt <- mat[l1, l2]

# Filter metadata to contain same samples as mat
meta.flt <- meta.lsc %>%
  distinct(Sample, .keep_all = TRUE) %>%
  # Remove NAs
  filter(!is.na(TP53)) %>%
  column_to_rownames('Sample')

# Keep only rows present in mat
meta.flt <- meta.flt[rownames(meta.flt) %in% rownames(mat.flt), ]

col = c("snv" = " black", "loss" = "darkblue", "gain" = "darkred", "other" = 'darkorange')

alter_fun = list(
    background = alter_graphic("rect", fill = "lightgrey"),   
    snv = alter_graphic("rect", fill = 'black', height = 0.5, width = 0.6),
    loss = alter_graphic("rect", fill = 'darkblue', height = 0.5, width = 0.6),
    gain = alter_graphic("rect", fill = 'darkred', height = 0.5, width = 0.6),
    other = alter_graphic("rect", fill = 'darkorange', height = 0.5, width = 0.6)
)

column_title = paste0("OncoPrint for AML samples (n = ", nrow(mat), ")")
heatmap_legend_param = list(title = "Alternations", at = c("snv", "other", "loss", "gain"), 
                            labels = c("SNV", "Other", "Deletion", "Duplication"),
                            title_gp = gpar(fontsize = 6),
                            labels_gp = gpar(fontsize = 6)
)

# With Mono
col.cell.types<- c('HSC' = '#A7303099',
                   'LMPP' = '#CD534C99',
                   'MEP' = '#8F770099',
                   'MoDe' = 'darkred')

column_ha <- HeatmapAnnotation(cbar = anno_oncoprint_barplot(axis_param = list(gp = gpar(fontsize = 6))
                                                             ),
                               "LSC Type" = meta.flt$LSC_type,
                               col = list("LSC Type" = col.cell.types),
                               annotation_name_gp = gpar(fontsize = 6),
                               annotation_legend_param = list(
                                 "LSC Type" = list(
                                   title_gp = gpar(fontsize = 6),
                                   labels_gp = gpar(fontsize = 6))
                               ),
                               height = unit(3, "cm"))

column_ha <- re_size(column_ha, annotation_height = c(unit(1, "cm"), unit(0.25, "cm")), height = unit(1.25, "cm"))

ht <- oncoPrint(t(mat.flt), alter_fun = alter_fun, col = col, column_title = NULL,
               top_annotation = column_ha,
               column_names_gp = gpar(fontsize = 8),
               pct_gp = gpar(fontsize = 8),
               # Split column based on celltype
               column_split = meta.flt$LSC_type,
               heatmap_legend_param = heatmap_legend_param)

ht

# Save without histogram (barplot) on the right side
ht_minimal <- oncoPrint(t(mat.flt), alter_fun = alter_fun, col = col, column_title = NULL,
               top_annotation = column_ha,
               right_annotation = NULL,
               column_names_gp = gpar(fontsize = 6),
               pct_gp = gpar(fontsize = 6),
               # Split column based on celltype
               column_split = meta.flt$LSC_type,
               heatmap_legend_param = heatmap_legend_param)


pdf(paste0(output_dir, 'Combined_cohort_oncoPrint_LSCtypes.pdf'), width = 4.5, height = 3)
ht_minimal
dev.off()
```

