---
title: "R Notebook - Supplementary Figure 3"
output:
  github_document:
    toc: true
  html_document:
    df_print: paged
    keep_md: true
---
This notebook covers plots in Supplementary Figure 3.

```{r setup, include=FALSE}
    knitr::opts_knit$set(root.dir = normalizePath('/Users/leppam/dkfz/venetoclax/manuscripts/code_github/')) 
```

Required packages and directories.

```{r, message=F}
library(data.table)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(forcats)
library(ggplot2)
library(ggbeeswarm)
library(cowplot)
library(readxl)
library(broom)

input_dir <- "./input/"
output_dir <- "./output/"

```

Read in BEAT AML data.

```{r}
# Normalized expression data
expr <- fread(paste0(input_dir, 'beataml_waves1to4_norm_exp_dbgap.txt'))

# Clinical data
df.clinical <- read_excel(paste0(input_dir, 'beataml_wv1to4_clinical.xlsx'), sheet = 'summary', col_types = c('text'))

# Normalized cibersortx deconvolution data
beat_deconvolution <- fread(paste0(input_dir, 'beat_deconvolution.tsv'), sep = "\t")

# Drug sensitivity data from samples with deconvolution data
drug_responses <- fread(paste0(input_dir, 'beat_drug_responses.tsv'), sep = "\t")

# Drug families
df_drug.families <- fread(paste0(input_dir, 'beataml_drug_families.tsv'))

```

# Supplementary Figure 3A

Plot LAMP5 expression in BEAT AML.

```{r, fig.height = 8, fig.width = 8}

# Pattern to detect amplification involving chromosome 8
amp_chr8_pattern <- "(\\+8|dup\\(8|dup\\(8[pq]|amp\\(8|8[pq]\\+|i\\(8q\\)|hsr\\(8|der\\(8|mar\\(.*8)"
# Pattern to detect TP53 deletion
tp53_del_pattern <- "del\\(17p|17p-|-17|del\\(17\\)\\([^\\)]*p13(?:[.0-9]*)?\\)"
# Pattern for detect t(8;16) 
trans_chr8_chr16_pattern <- "t\\(8\\;16"
# Pattern to detect Ras pathway mutations
ras_pathway_genes <- c("KRAS", "NRAS", "HRAS","BRAF","NF1", "PTPN11")
ras_pattern <- paste0("(?i)\\b(", paste(ras_pathway_genes, collapse = "|"), ")\\b")
# Pattern to detect splicing factor mutations
splice_genes <- c("SF3B1", "SRSF2", "U2AF1", "ZRSR2")
splice_pattern <- str_c("\\b(", str_c(splice_genes, collapse = "|"), ")\\b")

# Create df for LAMP5 expression with clinical data
pos.tail <- 3.5

expr.lamp5 <- expr %>%
  filter(display_label == 'LAMP5') %>%
  pivot_longer(cols = where(is.numeric), names_to = 'dbgap_sample_id', values_to = 'expr') %>%
  mutate(group = case_when(expr > pos.tail ~ 'High',
                                 TRUE ~ 'Low')) %>%
  # add clinical data
  left_join(., df.clinical, by = c('dbgap_sample_id' = 'dbgap_rnaseq_sample')) %>%
  # add information about chromosome 8 amplifications
  mutate('Dup8' = case_when(str_detect(karyotype, amp_chr8_pattern) ~ 'Yes',
                              is.na(karyotype) ~ 'Unknown',
                              TRUE ~ 'No')) %>%
  # add information about t(8;16) status
  mutate('t(8;16)' = case_when(str_detect(karyotype, trans_chr8_chr16_pattern) ~ 'Yes',
                              is.na(karyotype) ~ 'Unknown',
                              TRUE ~ 'No')) %>%
  # add information about KMT2A-re status
  mutate('KMT2A-re' = case_when(str_detect(consensusAMLFusions, 'KMT2A_re') ~ 'Yes',
                               consensusAMLFusions == ' N/A' ~ 'Unknown',
                               TRUE ~ 'No')) %>%
  # add information about tp53 status
  mutate('-17/-17p13/TP53' = case_when(str_detect(karyotype, tp53_del_pattern) | str_detect(variantSummary, 'TP53') ~ 'Yes',
                                 is.na(karyotype) | is.na(variantSummary) ~ 'Unknown',
                                 TRUE ~ 'No')) %>%
  # add information about common SNVs
  mutate('IDH1/2' = case_when(str_detect(variantSummary, 'IDH1|IDH2') ~ 'Yes',
                                is.na(variantSummary) ~ 'Unknown',
                                TRUE ~ 'No'),
         ASXL1 = case_when(str_detect(variantSummary, 'ASXL1') ~ 'Yes',
                               is.na(variantSummary) ~ 'Unknown',
                               TRUE ~ 'No'),
         RUNX1 = case_when(str_detect(variantSummary, 'RUNX1') ~ 'Yes',
                               is.na(variantSummary) ~ 'Unknown',
                               TRUE ~ 'No'),
         DNMT3A = case_when(str_detect(variantSummary, 'DNMT3A') ~ 'Yes',
                               is.na(variantSummary) ~ 'Unknown',
                               TRUE ~ 'No'),
         TET2 = case_when(str_detect(variantSummary, 'TET2') ~ 'Yes',
                               is.na(variantSummary) ~ 'Unknown',
                               TRUE ~ 'No'),
         'FLT3-ITD' = case_when(str_detect(variantSummary, 'FLT3-ITD') ~ 'Yes',
                               is.na(variantSummary) ~ 'Unknown',
                               TRUE ~ 'No'),
         NPM1 = case_when(str_detect(variantSummary, 'NPM1') ~ 'Yes',
                                  is.na(variantSummary) ~ 'Unknown',
                                  TRUE ~ 'No'),
         SF = case_when(str_detect(variantSummary, splice_pattern) ~ 'Yes',
                             is.na(variantSummary) ~ 'Unknown',
                             TRUE ~ 'No'),
         RAS = case_when(str_detect(variantSummary, ras_pattern) ~ 'Yes',
                             is.na(variantSummary) ~ 'Unknown',
                             TRUE ~ 'No')) %>%
  # correct typos inconsistenceis in clinical columns
  mutate(specificDxAtAcquisition = case_when(specificDxAtAcquisition %in% c('unknown', 'Unknown') ~ 'Unknown',
                                             TRUE ~ specificDxAtAcquisition)) %>%
  # only keep sample with clinical data
  filter(!is.na(dbgap_subject_id))


# Plot LAMP5 expression in disease subgroups
# Order specificDxAtAcquisition based on LAMP5 expression
group.order <- expr.lamp5 %>%
  group_by(specificDxAtAcquisition) %>%
  summarise(mean_expr = mean(expr)) %>%
  arrange(desc(mean_expr)) %>%
  pull(specificDxAtAcquisition)

p_lamp5 <- expr.lamp5 %>%
  ggplot(aes(x = factor(specificDxAtAcquisition, levels = group.order), y = expr)) +
  ggbeeswarm::geom_quasirandom(aes(fill = specificDxAtAcquisition), shape = 21, color = 'black') +
  geom_hline(yintercept = pos.tail, linetype = 'dashed') +
  stat_summary(fun.data = "mean_cl_boot", colour = "darkred", linewidth = 0.5, size = 0.5) +
  labs(
    title = "LAMP5 expression in disease subgroups",
    x = NULL,
    y = "Normalized RPKM"
  ) +
  cowplot::theme_cowplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = 'none')

p_lamp5  

save_plot(paste0(output_dir, 'LAMP5_expr_beat_aml_disease_subgroups.pdf'), p_lamp5, base_height = 8, base_width = 8)

```
# Supplementary Figure 3B

Plot multivariate linear regression of LAMP5 and genetics.

```{r}
# Variables to include
variables <- c('consensus_sex','ageAtSpecimenAcquisition',
              'Dup8','t(8;16)','-17/-17p13/TP53', 'KMT2A-re',
              'IDH1/2','ASXL1','RUNX1','DNMT3A','TET2','FLT3-ITD','NPM1','SF','RAS')

# To avoid issues with special characters
variables_escaped <- str_c("`", variables, "`")

variables_formula <- str_c(variables_escaped, collapse = "+")
full_formula <- as.formula(paste("expr ~", variables_formula))

# Multiple Linear Regression
# Remove rows that have NAs for the chosen variables
expr.lamp5.flt <- expr.lamp5 %>%
  mutate(ageAtSpecimenAcquisition = as.numeric(ageAtSpecimenAcquisition)) %>%
  select(all_of(variables), expr) %>%
  drop_na() %>%
  # Remove all "Unknown" rows
  filter(!if_any(everything(), ~ . == "Unknown"))

output <- lm(full_formula, data = expr.lamp5.flt)

summary_dfs <- summary(output)

# Get tidy results
lmod.tidy <- tidy(output, exponentiate = TRUE, conf.int = TRUE) %>%
  rename(variable = term, pval = p.value, CI_low = conf.low, CI_high = conf.high) %>%
  filter(variable != '(Intercept)') %>%
  mutate(label = case_when(pval < 0.001 ~ "***",
                           pval < 0.01 ~ "**",
                           pval < 0.05 ~ "*",
                           TRUE ~ '')) %>%
  # clean up variable column
  mutate(variable = str_remove_all(variable, '`')) %>%
  mutate(variable = str_remove(variable, 'Yes')) %>%
  mutate(variable = str_remove(variable, 'consensus_sex')) %>%
  mutate(variable = str_replace(variable, 'ageAtSpecimenAcquisition', 'Age'))

# Plot Beta Estimate
p_reg <- lmod.tidy %>%
  ggplot(aes(x = reorder(variable, estimate), y = estimate)) +
  geom_point(data = . %>% filter(pval > 0.05), size = 3) +
  geom_point(data = . %>% filter(pval < 0.05), color = 'darkred', size = 3) +
  geom_errorbar(data = . %>% filter(pval > 0.05), aes(ymin = CI_low, ymax = CI_high), width = 0.2) +
  geom_errorbar(data = . %>% filter(pval < 0.05), aes(ymin = CI_low, ymax = CI_high), color = 'darkred', width = 0.2) +
  geom_text(aes(label = label), hjust = 0.5, vjust = -0.3) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray") +
  scale_y_log10(breaks = c(0.1, 1, 10, 100, 1e3, 1e4, 1e5, 1e6, 1e7)) +
  scale_x_discrete(expand = expansion(0.06,0)) +
  coord_flip() +
  cowplot::theme_cowplot() +
  labs(
    subtitle = "Linear Model Coefficients",
    title = 'LAMP5 expression',
    x = "",
    y = "Beta Estimate (95% CI)"
  ) +
  theme(axis.text.x=element_text(angle=45, hjust=0.5, vjust = 0.5))

p_reg

save_plot(paste0(output_dir, 'LAMP5_expr_beat_aml_linear_regression.pdf'), p_reg, base_height = 5, base_width = 5)

```

# Supplementary Figure 3D

Plot CIBERSORTx deconvolution and ex vivo drug response results.

Ex vivo drug response in AML patient samples from BEAT-AML (Tyler et al. 2018) was measured through the AUC metric where low AUC score represents sensitivity and high score resistance. Acute promyelocytic leukaemia samples were excluded from the analysis. AUC values were scaled and multiplied by −1 to represent sensitivity in each treatment condition. Pearson correlation was used to measure association between cell type abundance and drug sensitivities. 

```{r}

# Function to plot results
drug_obs_correlations <- function(drug_responses, beat_deconvolution, celltype) {
  print(celltype)
  # Make sure rownames are the same between the data frames
  drug_responses <- drug_responses[rownames(beat_deconvolution), ]
  print(dim(drug_responses))
  all(rownames(drug_responses) == rownames(beat_deconvolution))
  
  correlations <- data.frame(pearson = numeric(), pvalue = numeric(), Drug = character(), row.names = character(0))

  for (drug in colnames(drug_responses)) {
    temp <- data.frame(score = beat_deconvolution[, celltype], auc = -as.vector(drug_responses[, drug]))
    temp <- temp[complete.cases(temp), ]
    pearson_result <- cor.test(temp$score, temp$auc, method = "pearson")
    data <- data.frame(pearson = pearson_result$estimate, pvalue = pearson_result$p.value, Drug = drug, row.names = drug)
    correlations <- rbind(correlations, data)
  }

  rownames(correlations) <- NULL
  return(correlations)
}

# Determine correlations between cell types and drug sensitivity (AUC) in all samples.

drug_responses <- drug_responses %>%
  column_to_rownames('patient')

beat_deconvolution <- beat_deconvolution %>%
  column_to_rownames('patient')

results <- data.frame()

celltypes <- c("HSC","LMPP","MEP", "GMP","Mono","ProgDC","pDC","cDC2")

for (celltype in celltypes) {
  temp <- drug_obs_correlations(drug_responses, beat_deconvolution, celltype)
  temp$CellType <- celltype
  results <- rbind(results, temp)
}

df.correlation <- results

# Perform multiple testing correction and identify significant cell - drug relationships
df.correlation$FDR <- p.adjust(df.correlation$pvalue, method = "fdr")
# significantly sensitive at pval < 0.05 and abs(pearson) > 0.3
sensitive_drugs <- unique(df.correlation$Drug[df.correlation$pvalue < 0.05 & abs(df.correlation$pearson) > 0.3])

df.correlation$logFDR <- log10(df.correlation$FDR)
df.correlation$CellType <- factor(df.correlation$CellType, levels = c("HSC","LMPP","MEP", "GMP","Mono","ProgDC","pDC","cDC2"))

# Modify tables for plotting
df_pearson <- df.correlation %>% 
  filter(Drug %in% sensitive_drugs) %>%
  select(pearson, CellType, Drug) %>%
  pivot_wider(names_from = Drug, values_from = pearson) %>%
  column_to_rownames('CellType')

df_FDR <- df.correlation %>% 
  filter(Drug %in% sensitive_drugs) %>%
  select(FDR, CellType, Drug) %>%
  pivot_wider(names_from = Drug, values_from = FDR) %>%
  column_to_rownames('CellType')

celltypes <- c("HSC","LMPP","MEP", "GMP","Mono","ProgDC","pDC","cDC2")

family.order <- c("RAF_MEK_ERK","MEK","HDAC","RTK_TIE","RTK_RET","RTK_VEGFRs","RTK_ALK_MET","RTK_TAM","RTK_ERBB","RTK_TYPE_III","BTK_TEC","RTK_INSR_IGF1R","CDK4_6","pan_CDK","PKC","IkK","FAKs","AURK","BCL2","Other")

# Wrangle data
df_plot <- df.correlation %>% 
  filter(Drug %in% sensitive_drugs) %>%
  select(FDR, pearson, CellType, Drug) %>%
  # Add drug family
  left_join(., df_drug.families, by = c('Drug' = 'inhibitor'), multiple = 'first') %>%
  mutate(family = case_when(is.na(family) ~ 'Other',
                            TRUE ~ family)) %>%
  mutate(family = factor(family, levels = family.order),
         CellType = factor(CellType, levels = celltypes),
         Drug = str_remove(Drug, ' .*')
         ) %>%
  mutate(CellType = factor(CellType, levels = celltypes)) %>%
  mutate(FDR_label = case_when(FDR < 0.05 ~ '*',
                               TRUE ~ '')
  )

# Plot
p_corr <- df_plot %>%
  mutate(label = paste0(Drug, "|", family)) %>%
  mutate(label = fct_reorder(label, as.integer(family) * 1000 + as.integer(as.factor(Drug)) )) %>%
  # Discretize FDR
  mutate(FDR_group = santoku::chop(FDR, breaks = c(0, 0.01,0.05,1))) %>%
  ggplot(aes(x = label, y = fct_rev(CellType), size = fct_rev(FDR_group), fill = pearson, label = FDR_label)) +
  geom_point(shape = 21) +
  #geom_text(size = 5, nudge_y = -0.15) +
  #viridis::scale_fill_viridis(option="inferno") +
  scale_fill_gradient2(low = scales::muted('blue'), mid = munsell::mnsl("N 8/0"), high = scales::muted('red'), space = "Lab") +
  scale_x_discrete(position = "top") +
  labs(x = NULL,
       y = NULL,
       title = 'BEAT ex vivo response deconvoluted',
       size = 'FDR',
       fill = 'cor(abundance, sensitivity)') +
  guides(x = legendry::guide_axis_nested(key = legendry::key_range_auto(sep = "\\|"), angle = 90, pad_discrete = 0.2)) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.key.size = unit(0.4, "cm"))

p_corr

save_plot(paste0(output_dir, 'Exvivo_response_deconvolution.pdf'), p_corr, base_height = 5, base_width = 8)

```

